---
title: "metaforExamples"
author: "Bo Wang"
date: "10/4/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressMessages(library(metafor))
```

### Examples from metafor-project.org

### Miller (1978) 
Create raw data and calculate transformed yi and vi.
```{r}
dat <- data.frame(xi=c(3, 6, 10, 1), ni=c(11, 17, 21, 6))
dat$pi <- with(dat, xi/ni)
dat <- escalc(measure="PFT", xi=xi, ni=ni, data=dat)
dat
#?escalc
```

Back transformation to raw proportions.
```{r}
transf.ipft(dat$yi, dat$ni)
#?trans.ipft
```


#### Meta-analysis of transformed proportions. 

1) Unweighted: the same as arithmetic mean of yi.
```{r}
res <- rma(yi, vi, method="FE", data=dat, weighted=FALSE)
res

coef(res)
mean(dat$yi)
```

Back-transform the estimate to linear scale. 

To back-transform the average, a value for the sample size is needed. Miller suggests to use the harmonic mean of the individual sample sizes in the inversion formula.

```{r}
predict(res, transf=transf.ipft.hm, targs=list(ni=dat$ni))
```

Therefore, the estimated true proportion based on the 4 studies is 0.32 (with 95% CI: 0.18 to 0.47).

# Question: how is the variance of back-transformed? Mathematically?


2) Weighted: by weights = inverse of variance.

# Question: why need to be homogeneous? What's Q(3) and p?

Since the true proportions appear to be homogeneous (e.g., Q(3)=2.18, p=0.54), a more efficient estimate of the true proportion can be obtained by using inverse-variance weights.

```{r}
res2 <- rma(yi, vi, method="FE", data=dat, weighted=TRUE)
res2

coef(res2)
sum(dat$yi*(1/dat$vi))/sum(1/dat$vi)
```

Back-transform the estimated transformed average.
```{r}
predict(res2, transf=transf.ipft.hm, targs=list(ni=dat$ni))
```
Therefore, the estimated true proportion is now equal to 0.36 (with 95% CI: 0.23 to 0.50).

#### Proportions equal to 0 or 1. 

When proportion=0 or 1, the variance of the untransformed proportion=np(1-p)=0, then the weight=1/variance=inf. If use logit transformation, logit(proportion)=log(proportion/1-proportion)=log(inf/-inf). Normal solution: add 0.5 to n of event. Fine if there are just a few extreme cases, but if the event is rare/common in general, need Freeman-Tukey transformation. 

```{r}
dat <- data.frame(xi=c(0,10), ni=c(10,10))
dat$pi <- with(dat, xi/ni)
##add=0 to specify no correction
dat <- escalc(measure="PFT", xi=xi, ni=ni, data=dat, add=0)
#dat2 <- escalc(measure="PFT", xi=xi, ni=ni, data=dat)
dat
#dat2
```
Back transformation.
```{r}
transf.ipft(dat$yi, dat$ni)
```


#### Forest plot

Create some data and calculated Freeman-Tukey transformed proportion and variance.
```{r}
dat <- data.frame(xi = c( 0,  4,  9, 16, 20),
                  ni = c(10, 10, 15, 20, 20))
dat$pi <- with(dat, xi/ni)
dat <- escalc(measure="PFT", xi=xi, ni=ni, data=dat, add=0)
dat
```

To obtain original proportions by back-transformation, we need individual study sample sizes.
```{r}
transf.ipft(dat$yi, dat$ni)
```

Calculate the weighted average and back transform the weighted average using the harmonic mean of the study sample sizes.
```{r}
res<-rma(yi, vi, method="FE", data=dat)
res

pred<-predict(res, transf=transf.ipft.hm, targs=list(ni=dat$ni))
pred
```

So the back-transformed estimated weighted average is 0.689 with CI (0.573, 0,794).

If we now want to draw a forest plot that includes not only the individual back-transformed proportions, but also the back-transformed average, things get complicated. If we use the following code, the resulting forest plot also uses the harmonic mean of the sample sizes for the back-transformation of the individual transformed proportions, which is not right.

```{r}
forest(res, transf=transf.ipft.hm, targs=list(ni=dat$ni), alim=c(0,1), refline=NA, digits=3)
```

In stead, we need to first obtain the CI bounds of the individual studies.
```{r}
dat.back <- summary(dat, transf=transf.ipft, ni=dat$ni)
dat.back
```

Now the back-transformation is applied to each transformed proportion with the study-specific sample sizes. The yi values are now the back-transformed values (i.e., the raw proportions) and the ci.lb and ci.ub values are the back-transformed 95% CI bounds.1)

Finally, we can create the forest plot by directly passing the observed outcomes (i.e., proportions) and the CI bounds to the function. Then the back-transformed average with the corresponding CI bounds obtained earlier can be added to the plot with the addpoly() function. We add a couple tweaks to make the final forest plot look nice:

```{r}
forest(dat.back$yi, ci.lb=dat.back$ci.lb, ci.ub=dat.back$ci.ub, psize=1,
       xlim=c(-0.5,1.8), alim=c(0,1), ylim=c(-1,8), refline=NA, digits=3, xlab="Proportion")
addpoly(pred$pred, ci.lb=pred$ci.lb, ci.ub=pred$ci.ub, row=-0.5, digits=3, mlab="FE Model", efac=1.3)
abline(h=0.5)
text(-0.5, 7, "Study",               pos=4)
text( 1.8, 7, "Proportion [95% CI]", pos=2)
```



#### 2 X 2 tables
```{r}
dat <- data.frame(
age = c("Age <55", "Age 55+"),
ai = c(8,22),
bi = c(98,76),
ci = c(5,16),
di = c(115,69))
dat
```

Stratum-Specific and Crude Risk Differences

The stratum-specific risk differences can be computed with:
```{r}
summary(escalc(ai=ai, bi=bi, ci=ci, di=di, data=dat, measure="RD", digits=3, append=FALSE))
```

On the other hand, the crude risk difference for these data can be obtained with:
```{r}
summary(escalc(ai=sum(ai), bi=sum(bi), ci=sum(ci), di=sum(di), data=dat, measure="RD", digits=3, append=FALSE))
```



