---
title: "Methods"
author: "Bo Wang"
date: "10/4/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(metafor))
```

```{r}
#sample data for proportions (from metafor package see doc of "escalc")
#dat <- dat.pritz1997
dat <- data.frame(xi=c(3, 6, 10, 0), ni=c(11, 17, 21, 6))
dat$pi <- with(dat, xi/ni)
dat

write.csv(dat, file = "dat.csv")

#sample data for rates (from metafor-project.org: http://www.metafor-project.org/doku.php/analyses:stijnen2010)
datr <- data.frame(
         age = c("35-44", "45-54", "55-64", "65-74", "75-84"),
         x1i = c(32, 104, 206, 186, 102),
         t1i = c(52407, 43248, 28612, 12663, 5317) / 10000)#,
         # x2i = c(2, 12, 28, 28, 31),
         # t2i = c(18790, 10673, 5710, 2585, 1462) / 10000)
datr
```

1. Discrete likelihood methods, which model the proportion of events or the incidence rate in a study using the binomial or Poisson distribution, respectively. These are theoretically motivated choices.

```{r}
#reference:http://www.metafor-project.org/doku.php/analyses:stijnen2010

#proportions: binomial-normal 
#fixed effect
res.bin.fe <- rma.glmm(measure="PLO", method="FE", xi=xi, ni=ni, data=dat)
res.bin.fe
predict(res.bin.fe, transf=transf.ilogit)

#forest plot
forest(res.bin.fe)
forest(res.bin.fe, transf=transf.ilogit)

#first calculate yi and SEi: won't work
#dat.bin.fe <- escalc(measure="PLO", xi=xi, ni=ni, data=dat)
#res.bin.fe.alt <- rma.glmm(method="FE", yi=yi, vi=vi, data=dat.bin.fe, measure="PLO")

#random effect
res.bin.ml <- rma.glmm(measure="PLO", method="ML", xi=xi, ni=ni, data=dat)
res.bin.ml
predict(res.bin.ml, transf=transf.ilogit)
```

```{r}
#rates: poisson-normal 
#fixed effect
res.pois.fe <- rma.glmm(measure="IRLN", method="FE", xi=x1i, ti=t1i, data=datr)
res.pois.fe
predict(res.pois.fe, transf=exp)

#random effect
res.pois.ml <- rma.glmm(measure="IRLN", method="ML", xi=x1i, ti=t1i, data=datr)
res.pois.ml
predict(res.pois.ml, transf=exp)
```


2. Approximate methods, which approximate within-study variability with a normal distribution. Of the many variants that have been used, the three that are most interestinga use a normal distribution to approximate the distribution of:

a. Untransformed proportions or rates.
```{r}
#proportions:
#no transformation (with correction factor)
#note that yi =/= pi when pi = 1 b/c correction
dat.raw <- escalc(measure="PR", xi=xi, ni=ni, data=dat)
dat.raw

#fixed effect: inverse variance
res.raw.fe<-rma(yi, vi, method="FE", data=dat.raw, weighted=TRUE)
res.raw.fe
c(pred=coef(res.raw.fe), ci.lb=res.raw.fe$ci.lb, ci.up=res.raw.fe$ci.ub)

#random effect: DL
res.raw.dl<-rma(yi, vi, method="DL", data=dat.raw)
res.raw.dl
c(pred=coef(res.raw.dl), ci.lb=res.raw.dl$ci.lb, ci.up=res.raw.dl$ci.ub)
```

```{r}
#rates:
#no transformation 
datr.raw <- escalc(measure="IR", xi=x1i, ti=t1i, data=datr)
datr.raw

#fixed effect: inverse variance
resr.raw.fe<-rma(yi, vi, method="FE", data=datr.raw, weighted=TRUE)
resr.raw.fe
c(pred=coef(resr.raw.fe), ci.lb=resr.raw.fe$ci.lb, ci.up=resr.raw.fe$ci.ub)

#random effect: DL
resr.raw.dl<-rma(yi, vi, method="DL", data=datr.raw)
resr.raw.dl
c(pred=coef(resr.raw.dl), ci.lb=resr.raw.dl$ci.lb, ci.up=resr.raw.dl$ci.ub)
```

b. Canonical transformations for proportions (logit transformations) and rates (logarithmic transformations).
```{r}
#proportions:
#logit transformation (with correction factor)
dat.logit <- escalc(measure="PLO", xi=xi, ni=ni, data=dat)
dat.logit

#back-transformation of individual estimates
#note that the back-transformed individual proportion =/= pi when pi = 1 b/c correction
transf.ilogit(dat.logit$yi, dat.logit$ni)

#fixed effect: inverse variance
res.logit.fe<-rma(yi, vi, method="FE", data=dat.logit, weighted=TRUE)
res.logit.fe

#back-transformation of average estimate: FE
predict(res.logit.fe, transf=transf.ilogit)

#random effect: DL
res.logit.dl<-rma(yi, vi, method="DL", data=dat.logit)
res.logit.dl

#back-transformation of average estimate: DL
predict(res.logit.dl, transf=transf.ilogit)
```

```{r}
#rates:
#log transformation 
datr.log <- escalc(measure="IRLN", xi=x1i, ti=t1i, data=datr)
datr.log

#same as
yi.log<-log(datr$x1i/datr$t1i)
yi.log

#fixed effect: inverse variance
res.log.fe<-rma(yi, vi, method="FE", data=datr.log, weighted=TRUE)
res.log.fe

#back-transformation of average estimate: FE
predict(res.log.fe, transf=exp)

#random effect: DL
res.log.dl<-rma(yi, vi, method="DL", data=datr.log)
res.log.dl

#back-transformation of average estimate: DL
predict(res.log.dl, transf=exp)
```

c. Variance stabilizing transformations for proportions (arcsine transformations) and rates (square root transformations).
```{r}
#proportions:
#arcsine square root transformation (without correction factor)
dat.as <- escalc(measure="PAS", xi=xi, ni=ni, data=dat, add=0)
dat.as

#back-transformation of individual estimates: matched pi
transf.iarcsin(dat.as$yi, dat.as$ni)

#fixed effect: inverse variance
res.as.fe<-rma(yi, vi, method="FE", data=dat.as, weighted=TRUE)
res.as.fe

#back-transformation of average estimate: FE
predict(res.as.fe, transf=transf.iarcsin)

#random effect: DL
res.as.dl<-rma(yi, vi, method="DL", data=dat.as)
res.as.dl

#back-transformation of average estimate: DL
predict(res.as.dl, transf=transf.iarcsin)
```

```{r}
#freeman-tukey (without correction factor)
dat.ft <- escalc(measure="PFT", xi=xi, ni=ni, data=dat, add=0)
dat.ft

#fixed effect
res.ft.fe <- rma(yi, vi, method="FE", data=dat.ft)
predict(res.ft.fe, transf=transf.ipft.hm, targs=list(ni=dat.ft$ni))

#random effect
res.ft.dl <- rma(yi, vi, method="DL", data=dat.ft)
predict(res.ft.dl, transf=transf.ipft.hm, targs=list(ni=dat.ft$ni))
```

```{r}
#rates:
#square root transformation
datr.sqrt <- escalc(measure="IRS", xi=x1i, ti=t1i, data=datr)
datr.sqrt

#same as
sqrt(datr$x1i/datr$t1i)

#fixed effect: inverse variance
res.sqrt.fe<-rma(yi, vi, method="FE", data=datr.sqrt, weighted=TRUE)
res.sqrt.fe

#back-transformation of average estimate: FE
predict(res.sqrt.fe, transf=transf.isqrt)

#random effect: DL
res.sqrt.dl<-rma(yi, vi, method="DL", data=datr.sqrt)
res.sqrt.dl

#back-transformation of average estimate: DL
predict(res.sqrt.dl, transf=transf.isqrt)
```

#### Compare analysis results
```{r}
#proportions
results.p<-round(rbind(as.numeric(as.character(unlist(predict(res.bin.fe, transf=transf.ilogit))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.bin.ml, transf=transf.ilogit))[c(1,3,4)])),
                       c(pred=coef(res.raw.fe), ci.lb=res.raw.fe$ci.lb, ci.up=res.raw.fe$ci.ub),
                       c(pred=coef(res.raw.dl), ci.lb=res.raw.dl$ci.lb, ci.up=res.raw.dl$ci.ub),
                       as.numeric(as.character(unlist(predict(res.logit.fe, transf=transf.ilogit))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.logit.dl, transf=transf.ilogit))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.as.fe, transf=transf.iarcsin))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.as.dl, transf=transf.iarcsin))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.ft.fe, transf=transf.ipft.hm, targs=list(ni=dat.ft$ni)))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.ft.dl, transf=transf.ipft.hm, targs=list(ni=dat.ft$ni)))[c(1,3,4)]))), 3)

rownames(results.p)<-c("Binom-normal FE logit", 
                       "Binom-normal ML logit",
                       "Untransformed FE",
                       "Untransformed ML",
                       "Logit FE",
                       "Logit DL",
                       "Arcsine FE",
                       "Arcsine DL",
                       "Freeman-Tukey FE",
                       "Freeman-Tukey DL")

results.p
```

```{r}
#rates
results.r<-round(rbind(as.numeric(as.character(unlist(predict(res.pois.fe, transf=exp))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.pois.ml, transf=exp))[c(1,3,4)])),
                       c(pred=coef(resr.raw.fe), ci.lb=resr.raw.fe$ci.lb, ci.up=resr.raw.fe$ci.ub),
                       c(pred=coef(resr.raw.dl), ci.lb=resr.raw.dl$ci.lb, ci.up=resr.raw.dl$ci.ub),
                       as.numeric(as.character(unlist(predict(res.log.fe, transf=exp))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.log.dl, transf=exp))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.sqrt.fe, transf=transf.isqrt))[c(1,3,4)])),
                       as.numeric(as.character(unlist(predict(res.sqrt.dl, transf=transf.isqrt))[c(1,3,4)]))
                       ), 3)

rownames(results.r)<-c("Pois-normal FE log", 
                       "Pois-normal ML log",
                       "Untransformed FE",
                       "Untransformed ML",
                       "Log FE",
                       "Log DL",
                       "Sqrt FE",
                       "Sqrt DL")

results.r
```




